From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-30 upi.tex
add a note regarding the categorical link.
minor tweaks to abstract and introduction.

\begin{abstract}
  We establish a close connection between a reversible programming language
  based on type isomorphisms and a formally presented univalent universe. The
  correspondence relates combinators witnessing type isomorphisms in the
  programming language to paths in the univalent universe and combinator
  optimizations in the programming language to 2-paths in the univalent
  universe. The result suggests a simple computational interpretation of paths
  and of univalence in terms of familiar programming constructs whenever the
  universe in question is computable.
\end{abstract}

@@

The proceedings of the 2012 Symposium on Principles of Programming
Languages~\cite{Field:2012:2103656} included two apparently unrelated papers:
\emph{Information Effects} by James and Sabry and \emph{Canonicity for
  2-dimensional type theory} by Licata and Harper. The first paper, motivated by
the physical nature of
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
proposed, among other results, a reversible language $\Pi$ in which every
program is a type isomorphism. The second paper, motivated by the connections
between homotopy theory and type theory~\cite{vv06,hottbook}, proposed a
judgmental formulation of intensional dependent type theory with a
twice-iterated identity type. During the presentations and ensuing discussions
at the conference, it became apparent, at an intuitive and informal level, that
the two papers had strong similarities. Formalizing the precise connection was
far from obvious, however.

@@

>>>>

\begin{abstract}
  We establish a close connection between a reversible programming language
  based on type isomorphisms and a formally presented univalent universe. The
  correspondence relates combinators witnessing type isomorphisms in the
  programming language to paths in the univalent universe; and combinator
  optimizations in the programming language to 2-paths in the univalent
  universe. The result suggests a simple computational interpretation of paths
  and of univalence in terms of familiar programming constructs whenever the
  universe in question is computable.
\end{abstract}

@@

The proceedings of the $2012$ Symposium on Principles of Programming
Languages~\cite{Field:2012:2103656} included two apparently unrelated papers:
\emph{Information Effects} by James and Sabry and \emph{Canonicity for
  2-dimensional type theory} by Licata and Harper. The first paper, motivated by
the physical nature of
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
proposed, among other results, a reversible language $\Pi$ in which every
program is a type isomorphism. The second paper, motivated by the connections
between homotopy theory and type theory~\cite{vv06,hottbook}, proposed a
judgmental formulation of intensional dependent type theory with a
twice-iterated identity type. During the presentations and ensuing discussions
at the conference, it became apparent, at an intuitive and informal level, that
the two papers had strong similarities. Formalizing the precise connection was
far from obvious, however.

@@

It is worthwhile mentioning that the above derivation could also be drawn as 
one (large!) commutative diagram in an appropriate category, with each
$\byisotwo{-}$ as a $2$-arrow (and representing a natural isomorphism).
See~\cite{Shulmann} for that interpretation.
